---
title: "RforMarineScience_Modulde2_Week3-4"
author: "Andreanne Paul-Chowdhury"
date: "2026-08-11"
output: html_document
---
# Workshop Overview

This workshop introduces foundational data-wrangling and visualisation techniques in R. The aim is to learn how ecological datasets can be inspected, filtered, transformed, summarised and visualised to identify meaningful biological patterns.

```{r}
#Workshop 1 - Foundations of Data Science

# Load packages
library(tidyverse)
library(palmerpenguins)

```

```{r}
# Exploring the Dataset
# Load the Palmer Penguins dataset
data("penguins")

# View the data structure
glimpse(penguins)

# View a statistical summary
summary(penguins)

## Selecting Variables
# Select morphological variables
morphology_metrics <- penguins |>
  select(
    species,
    island,
    bill_length_mm,
    bill_depth_mm,
    body_mass_g
  )

glimpse(morphology_metrics)

## Filtering Observations
# Keep only Adelie penguins
adelie_penguins <- penguins |>
  filter(species == "Adelie")

# Keep penguins heavier than 4500 g
heavy_penguins <- penguins |>
  filter(body_mass_g > 4500)

# Keep Gentoo penguins sampled on Biscoe Island
biscoe_gentoo <- penguins |>
  filter(
    species == "Gentoo",
    island == "Biscoe"
  )

## Arranging Observations
# Smallest penguins first
lightest_first <- penguins |>
  arrange(body_mass_g)

# Heaviest penguins first
heaviest_first <- penguins |>
  arrange(desc(body_mass_g))

head(heaviest_first)

## Creating New Variables
penguin_ratios <- penguins |>
  mutate(
    body_mass_kg = body_mass_g / 1000,
    bill_ratio = bill_length_mm / bill_depth_mm
  )

glimpse(penguin_ratios)

# Ecological Summaries
#penguins |>
#  group_by(species) |>
#  summarise(
#    mean_mass_g = mean(body_mass_g)
#  )
##NA values
penguins |>
  group_by(species) |>
  summarise(
    mean_mass_g = mean(body_mass_g, na.rm = TRUE)
  )

## Mean Body Mass by Species and Island
body_mass_summary <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_body_mass_g = mean(body_mass_g, na.rm = TRUE),
    .groups = "drop"
  )

body_mass_summary

# Data Visualisation
body_mass_plot <- penguins |>
  drop_na(body_mass_g) |>
  ggplot(
    aes(
      x = species,
      y = body_mass_g,
      fill = island
    )
  ) +
  geom_boxplot() +
  labs(
    title = "Penguin Body Mass by Species and Island",
    x = "Species",
    y = "Body mass (g)",
    fill = "Island"
  ) +
  theme_minimal()

body_mass_plot

#Save outputs
ggsave(
  filename = "outputs/workshop1_body_mass_plot.png",
  plot = body_mass_plot,
  width = 8,
  height = 6,
  dpi = 300
)

write_csv(
  body_mass_summary,
  "outputs/workshop1_body_mass_summary.csv"
)
```

```{r}
# Final Morphological Summary
mass_summary_final <- penguins |>
  group_by(species, island) |>
  summarise(
    n = sum(!is.na(body_mass_g)),
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),
    sd_mass_g = sd(body_mass_g, na.rm = TRUE),
    se_mass_g = sd_mass_g / sqrt(n),
    .groups = "drop"
  )

mass_summary_final

knitr::kable(
  mass_summary_final,
  digits = 2,
  caption = "Summary of penguin body mass by species and island."
)
```
```{r}
## Mean Body Mass and Variation
mass_compare_plot <- mass_summary_final |>
  ggplot(
    aes(
      x = species,
      y = mean_mass_g,
      colour = island
    )
  ) +
  geom_point(
    size = 3,
    position = position_dodge(width = 0.3)
  ) +
  geom_errorbar(
    aes(
      ymin = mean_mass_g - sd_mass_g,
      ymax = mean_mass_g + sd_mass_g
    ),
    width = 0.15,
    position = position_dodge(width = 0.3)
  ) +
  labs(
    title = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent ±1 standard deviation",
    x = "Species",
    y = "Mean body mass (g)",
    colour = "Island"
  ) +
  theme_minimal()

mass_compare_plot

## Comparing Standard Deviation and Standard Error
mass_se_plot <- mass_summary_final |>
  ggplot(
    aes(
      x = species,
      y = mean_mass_g,
      colour = island
    )
  ) +
  geom_point(
    size = 3,
    position = position_dodge(width = 0.3)
  ) +
  geom_errorbar(
    aes(
      ymin = mean_mass_g - se_mass_g,
      ymax = mean_mass_g + se_mass_g
    ),
    width = 0.15,
    position = position_dodge(width = 0.3)
  ) +
  labs(
    title = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent ±1 standard error",
    x = "Species",
    y = "Mean body mass (g)",
    colour = "Island"
  ) +
  theme_minimal()

mass_se_plot
```
### Interpretation
####The grouped analysis demonstrates how individual morphological observations can be transformed into population-level summaries. Comparing mean body mass and variation among species and islands makes differences between biological groups easier to identify than examining the raw observations alone. The exercise also demonstrates that the choice of uncertainty measure matters: standard deviation describes variation among individuals, whereas standard error describes the precision of the estimated group mean.

```{r}
ggsave(
  filename = "outputs/workshop1_mass_summary_plot.png",
  plot = mass_compare_plot,
  width = 8,
  height = 6,
  dpi = 300
)
write_csv(
  mass_summary_final,
  "outputs/workshop1_mass_summary.csv"
)
```


#Workshop 2
# Workshop Overview

This workshop develops advanced data-wrangling skills for working with complex and messy ecological datasets. The workflow focuses on tidy data, reshaping datasets, standardising text and dates, joining related datasets, managing missing values and explicit zeros, and integrating multiple sources of biological and environmental data.
```{r}
# Load packages
library(tidyverse)
library(palmerpenguins)
library(readxl)
library(lubridate)
library(here)

# Tidy Data
##tidy data aren't necessarily prettier, they're structured so R can analyse them efficiently.

# Reshaping Data

## Pivoting Longer
data("penguins")

penguins_long <- penguins |>
  pivot_longer(
    cols = c(
      bill_length_mm,
      bill_depth_mm,
      flipper_length_mm,
      body_mass_g
    ),
    names_to = "measurement_type",
    values_to = "value"
  )

glimpse(penguins_long)

penguins_long |>
  drop_na(value) |>
  ggplot(
    aes(
      x = value,
      fill = species
    )
  ) +
  geom_histogram(
    bins = 30,
    alpha = 0.7,
    colour = "black"
  ) +
  facet_wrap(
    ~ measurement_type,
    scales = "free_x"
  ) +
  theme_minimal() +
  labs(
    title = "Morphometric Distributions Across Penguin Species",
    x = "Measurement value",
    y = "Frequency",
    fill = "Species"
  )

## Pivoting Wider
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g),
    .groups = "drop"
  )

mass_summary

mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

mass_matrix

# Cleaning Text and Dates
messy_sites <- tibble(
  site_id = c(
    " Nelly Bay",
    "nelly_bay",
    "NELLY BAY",
    " Geoffrey_Bay ",
    "geoffrey bay"
  )
)

clean_sites <- messy_sites |>
  mutate(
    site_clean = str_to_lower(site_id),
    site_clean = str_trim(site_clean),
    site_clean = str_replace_all(
      site_clean,
      pattern = " ",
      replacement = "_"
    )
  )

clean_sites

date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

date_1
date_2

date_1 == date_2

sensor_data <- tibble(
  raw_time = c(
    "14-05-2026 08:30:00",
    "14-05-2026 08:45:00",
    "14-05-2026 09:00:00"
  ),
  temperature = c(
    24.5,
    24.6,
    24.4
  )
)

sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

sensor_clean

#Practise a relational join
# Relational Data and Joins
island_metadata <- tibble(
  island_name = c(
    " biscoe",
    "Dream ",
    "Torgersen"
  ),
  station_install = c(
    "15/01/2003",
    "22-03-2004",
    "05/11/2001"
  ),
  latitude = c(
    -64.81,
    -64.73,
    -64.76
  )
)

island_metadata
#Clean it
clean_metadata <- island_metadata |>
  mutate(
    island_name = str_trim(island_name),
    island_name = str_to_title(island_name),
    station_install = dmy(station_install)
  )

clean_metadata

#Join with penguins
penguins_spatial <- penguins |>
  left_join(
    clean_metadata,
    by = join_by(island == island_name)
  )

glimpse(penguins_spatial)

#Why anti_join() matters
penguins |>
  anti_join(
    clean_metadata,
    by = join_by(island == island_name)
  )

# Missing Values and Ecological Zeros
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)
)
logger_clean <- logger_data |>
  mutate(
    temp_c = na_if(temp_c, -999)
  )

logger_clean

#Understand ecological zeros
raw_catch <- tibble(
  site = c(
    "Reef_1",
    "Reef_1",
    "Reef_2"
  ),
  species = c(
    "Pmaculatus",
    "Pleopardus",
    "Pmaculatus"
  ),
  count = c(
    5,
    2,
    8
  )
)

full_catch_matrix <- raw_catch |>
  complete(
    site,
    species,
    fill = list(count = 0)
  )

full_catch_matrix
```

```{r}
#Data rescue mission

# Estuary Fish Survey Data Rescue Mission

## Inspecting the Raw Data
list.files("data")

catch_sheets <- excel_sheets(
  here("data", "estuary_catch_log.xlsx")
)

catch_sheets

catch_test <- read_excel(
  here("data", "estuary_catch_log.xlsx"),
  sheet = catch_sheets[1]
)

glimpse(catch_test)
head(catch_test)

#Inspect all other 3 files
metadata_raw <- read_csv(
  here("data", "estuary_metadata.csv")
)

sonde_raw <- read_csv(
  here("data", "estuary_sonde_data.csv")
)

species_dictionary_raw <- read_csv(
  here("data", "species_dictionary.csv")
)

# Data Rescue Mission

## Checking Variable Names
# Catch data - first Excel sheet
names(catch_test)

# Site metadata
names(metadata_raw)

# Water-quality sonde data
names(sonde_raw)

# Species dictionary
#names(species_dictionary)

#Something was wrong dont know what but had to load file again
list.files("data")

species_dictionary_raw <- read_csv(
  here("data", "species_dictionary.csv")
)

names(species_dictionary_raw)
glimpse(species_dictionary_raw)

## Phase 1: Ingestion and Decontamination

### Combining Catch Records
catch_raw_all <- map_dfr(
  catch_sheets,
  ~ read_excel(
    here("data", "estuary_catch_log.xlsx"),
    sheet = .x
  )
)

glimpse(catch_raw_all)

nrow(catch_raw_all)

distinct(catch_raw_all, site)

#Standardise the catch data

catch_clean <- catch_raw_all |>
  mutate(
    # Clean site names
    site = str_squish(site),
    site = str_to_lower(site),
    site = str_replace_all(site, "[^a-z0-9]+", "_"),
    
    # Clean species names
    species = str_squish(species),
    species = str_to_lower(species),
    species = str_replace_all(species, "[^a-z0-9]+", "_")
  )

glimpse(catch_clean)

distinct(catch_clean, site)

distinct(catch_clean, species)

#Clean the site metadata
metadata_clean <- metadata_raw |>
  mutate(
    site_name = str_squish(site_name),
    site_name = str_to_lower(site_name),
    site_name = str_replace_all(
      site_name,
      "[^a-z0-9]+",
      "_"
    )
  ) |>
  rename(
    site = site_name
  )

metadata_clean

distinct(metadata_clean, site)

#Clean the species dictionary
species_dictionary_clean <- species_dictionary_raw |>
  mutate(
    common_name = str_squish(common_name),
    common_name = str_to_lower(common_name),
    common_name = str_replace_all(
      common_name,
      "[^a-z0-9]+",
      "_"
    )
  )

species_dictionary_clean

#Clean the sonde data
sonde_clean <- sonde_raw |>
  mutate(
    # Standardise site names
    site = str_squish(site),
    site = str_to_lower(site),
    site = str_replace_all(
      site,
      "[^a-z0-9]+",
      "_"
    ),
    
    # Convert timestamp text into real date-time data
    timestamp = dmy_hms(timestamp),
    
    # Convert sensor failure code to genuine missing data
    turbidity = na_if(turbidity, -999.0)
  )

glimpse(sonde_clean)

#Check the -999 gone 
sum(sonde_clean$turbidity == -999, na.rm = TRUE)
sum(is.na(sonde_clean$turbidity))

#Check if data sets will join
catch_clean |>
  distinct(site) |>
  anti_join(
    metadata_clean |>
      distinct(site),
    by = "site"
  )

## Phase 2: Relational Architecture

### Daily Water Quality Summary

sonde_daily <- sonde_clean |>
  mutate(
    date = as.Date(floor_date(timestamp, unit = "day"))
  ) |>
  group_by(site, date) |>
  summarise(
    mean_temperature = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    .groups = "drop"
  )

sonde_daily

glimpse(sonde_daily)

head(sonde_daily)

distinct(sonde_daily, site)

#Check catch date format
class(catch_clean$date)

class(sonde_daily$date)

catch_clean <- catch_clean |>
  mutate(
    date = as.Date(date)
  )
class(catch_clean$date)

#Translate common names to scientific names
catch_taxonomy <- catch_clean |>
  left_join(
    species_dictionary_clean,
    by = c("species" = "common_name")
  )

glimpse(catch_taxonomy)

distinct(
  catch_taxonomy,
  species,
  scientific_name
)

#Check that the toxonomy join worked
sum(is.na(catch_taxonomy$scientific_name))

#Join the spatial metadata
catch_spatial <- catch_taxonomy |>
  left_join(
    metadata_clean,
    by = "site"
  )

glimpse(catch_spatial)
sum(is.na(catch_spatial$zone))

#Join daily water-qualityd data
master_data <- catch_spatial |>
  left_join(
    sonde_daily,
    by = c("site", "date")
  )

glimpse(master_data)

#Check the Join
sum(is.na(master_data$zone))
sum(is.na(master_data$scientific_name))
sum(is.na(master_data$mean_salinity))
sum(is.na(master_data$mean_temperature))

#Use anti_join() if environmental values are missing
sum(is.na(master_data$mean_salinity))
catch_spatial |>
  distinct(site, date) |>
  anti_join(
    sonde_daily |>
      distinct(site, date),
    by = c("site", "date")
  )
#Inspect master dataset
head(master_data)
glimpse(master_data)
summary(master_data)
```

```{r}
# 1. Look at raw catch dates
head(catch_clean$date, 10)

# 2. Look at sonde daily dates
head(sonde_daily$date, 10)

# 3. Check the classes
class(catch_clean$date)
class(sonde_daily$date)

# 4. Check date ranges
range(catch_clean$date, na.rm = TRUE)
range(sonde_daily$date, na.rm = TRUE)

# 5. Make sure the sonde timestamp actually parsed correctly
sum(is.na(sonde_clean$timestamp))

# 6. See which site-date combinations fail to match
catch_spatial |>
  distinct(site, date) |>
  anti_join(
    sonde_daily |>
      distinct(site, date),
    by = c("site", "date")
  )
```

```{r}
head(sonde_raw$timestamp, 10)
#Correct sonde_clean
sonde_clean <- sonde_raw |>
  mutate(
    # Standardise site names
    site = str_squish(site),
    site = str_to_lower(site),
    site = str_replace_all(
      site,
      "[^a-z0-9]+",
      "_"
    ),

    # Correctly parse day/month/year hour:minute
    timestamp = dmy_hm(timestamp),

    # Convert turbidity sensor failure code to NA
    turbidity = na_if(turbidity, -999.0)
  )

sum(is.na(sonde_clean$timestamp))

head(sonde_clean$timestamp)

#recreate the daily sonde summary 
sonde_daily <- sonde_clean |>
  mutate(
    date = as.Date(timestamp)
  ) |>
  group_by(site, date) |>
  summarise(
    mean_temperature = mean(
      temperature,
      na.rm = TRUE
    ),
    mean_salinity = mean(
      salinity,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

sonde_daily

range(sonde_daily$date)

#Check that dates overlap
range(catch_clean$date)

range(sonde_daily$date)

catch_spatial |>
  distinct(site, date) |>
  anti_join(
    sonde_daily |>
      distinct(site, date),
    by = c("site", "date")
  )

#Rebuild master_data
master_data <- catch_spatial |>
  left_join(
    sonde_daily,
    by = c("site", "date")
  )
sum(is.na(master_data$mean_temperature))

sum(is.na(master_data$mean_salinity))

glimpse(master_data)
```

```{r}
## Phase 3: Zero-Catch Framework
catch_for_zero <- catch_taxonomy |>
  select(
    site,
    date,
    scientific_name,
    count
  ) |>
  group_by(
    site,
    date,
    scientific_name
  ) |>
  summarise(
    count = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

glimpse(catch_for_zero)

#Check for duplicates
catch_for_zero |>
  count(
    site,
    date,
    scientific_name
  ) |>
  filter(n > 1)

#How many observations?
nrow(catch_for_zero) #427
table(catch_for_zero$count == 0)

#Create the zero-catch framework
catch_complete <- catch_for_zero |>
  complete(
    nesting(site, date),
    scientific_name
  ) |>
  mutate(
    count = coalesce(count, 0)
  )

catch_complete

#See what changed
nrow(catch_for_zero)

nrow(catch_complete)

species_check <- catch_complete |>
  count(
    site,
    date,
    name = "n_species"
  )

species_check

distinct(
  species_check,
  n_species
)

#Reattach the spatial metadata
master_complete <- catch_complete |>
  left_join(
    metadata_clean,
    by = "site"
  )
glimpse(master_complete)

#Reattach the environmental data
master_complete <- master_complete |>
  left_join(
    sonde_daily,
    by = c("site", "date")
  )
glimpse(master_complete)

#Validate the finished master dataset
# No missing scientific names
sum(is.na(master_complete$scientific_name))
# No missing zones
sum(is.na(master_complete$zone))
# No missing salinity
sum(is.na(master_complete$mean_salinity))
# No missing temperature
sum(is.na(master_complete$mean_temperature))

#Check the number of zero catches
master_complete |>
  summarise(
    total_rows = n(),
    zero_catches = sum(count == 0),
    positive_catches = sum(count > 0)
  )
master_complete |>
  filter(site == "upper_ross") |>
  arrange(date, scientific_name) |>
  select(
    site,
    date,
    scientific_name,
    count
  ) |>
  head(12)

#Save
write_csv(
  master_complete,
  here(
    "outputs",
    "workshop2_clean_estuary_master.csv"
  )
)
```

```{r}
## Phase 4: Statistical Extraction
estuary_summary <- master_complete |>
  group_by(
    scientific_name,
    zone
  ) |>
  summarise(
    # Number of observations
    n = n(),

    # Fish abundance
    mean_count = mean(count, na.rm = TRUE),
    sd_count = sd(count, na.rm = TRUE),

    # Salinity
    avg_salinity = mean(mean_salinity, na.rm = TRUE),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),

    .groups = "drop"
  ) |>
  mutate(
    # Standard errors
    se_count = sd_count / sqrt(n),
    se_salinity = sd_salinity / sqrt(n)
  )

estuary_summary

glimpse(estuary_summary)
summary(estuary_summary)

#Round the values
estuary_portfolio_table <- estuary_summary |>
  select(
    scientific_name,
    zone,
    n,
    mean_count,
    sd_count,
    avg_salinity,
    sd_salinity
  ) |>
  mutate(
    across(
      c(
        mean_count,
        sd_count,
        avg_salinity,
        sd_salinity
      ),
      ~ round(.x, 2)
    )
  )

estuary_portfolio_table

knitr::kable(
  estuary_portfolio_table,
  caption = "Mean fish abundance and salinity by species and estuary zone."
)

```

```{r}
## Phase 5: Visual Communication
plot_data <- master_complete |>
  mutate(
    site = factor(
      site,
      levels = c(
        "upper_ross",
        "mid_ross",
        "lower_ross",
        "ross_mouth"
      )
    ),

    # Create labels that ggplot can display in italics
    facet_name = paste0(
      "italic('",
      scientific_name,
      "')"
    )
  )

levels(plot_data$site)

#Abundance-salinity figure
fish_salinity_plot <- plot_data |>
  ggplot(
    aes(
      x = mean_salinity,
      y = count,
      colour = site
    )
  ) +
  geom_point(
    alpha = 0.6,
    size = 2
  ) +
  facet_wrap(
    ~ facet_name,
    labeller = label_parsed,
    scales = "free_y"
  ) +
  labs(
    title = "Predatory Fish Abundance Along the Estuary Salinity Gradient",
    x = "Mean daily salinity",
    y = "Fish abundance (count)",
    colour = "Sampling site"
  ) +
  theme_minimal(base_size = 12)

fish_salinity_plot

#Add ecological trend line
fish_salinity_plot <- plot_data |>
  ggplot(
    aes(
      x = mean_salinity,
      y = count,
      colour = site
    )
  ) +
  geom_point(
    alpha = 0.55,
    size = 2
  ) +
  geom_smooth(
    aes(group = 1),
    method = "loess",
    se = FALSE,
    colour = "black",
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ facet_name,
    labeller = label_parsed,
    scales = "free_y"
  ) +
  labs(
    title = "Predatory Fish Abundance Along the Estuary Salinity Gradient",
    subtitle = "Points represent individual sampling events; lines show overall abundance trends",
    x = "Mean daily salinity",
    y = "Fish abundance (count)",
    colour = "Sampling site"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

fish_salinity_plot

#Save 
ggsave(
  filename = here(
    "outputs",
    "workshop2_fish_salinity_gradient.png"
  ),
  plot = fish_salinity_plot,
  width = 9,
  height = 6,
  dpi = 300
)
```

### Interpretation
The final visualisation integrates biological catch data with environmental measurements collected along the estuary. Faceting by species allows species-specific abundance patterns to be compared across the salinity gradient, while inclusion of explicit zero catches ensures that observed absences contribute to the ecological interpretation rather than being omitted from the analysis.

###Troubleshooting
During the workflow, all temperature and salinity values initially appeared as missing after joining the biological and environmental datasets. I traced the problem backwards through the pipeline and found that all 11,520 sonde timestamps had been converted to NA because the original timestamps contained hours and minutes but no seconds. Replacing dmy_hms() with the appropriate dmy_hm() parser corrected the timestamps and allowed the datasets to join successfully. This demonstrated the importance of validating intermediate outputs rather than assuming that code has worked simply because it runs without an error.