---
title: "RforMarineScience_Week3-4"
author: "Andreanne Paul-Chowdhury"
format: html
editor: visual
---

## Workshop 2: Advanced data wrangling: Extracting ecological signals from noisy systems

### 2.1 Introduction

Explain what was done in this workshop for readers of the code

### 2.2 Tidying data

How we make our dataset tidy is by following three interrelated rules. - Each variable must have its own column. -Each observation must have its own row. -Each value must have its own cell.

```{r}
library(tidyverse)
```

```{r}
table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```

```{r}
table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
```

```{r}
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583
```

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
#> # A tibble: 6 × 5
#>   country      year  cases population  rate
#>   <chr>       <int>  <int>      <int> <dbl>
#> 1 Afghanistan  1999    745   19987071 0.373
#> 2 Afghanistan  2000   2666   20595360 1.29 
#> 3 Brazil       1999  37737  172006362 2.19 
#> 4 Brazil       2000  80488  174504898 5.61 
#> 5 China        1999 212258 1272915272 1.67 
#> 6 China        2000 213766 1280428583 1.67

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```

### 2.3 Pivoting data to make it tidy

First step: understand what each variable and observation actually means. Second step:resolve 2 common problems with untidy data. -One variable is spread across multiple columns -One observation is scattered across multiple rows

```{r}
#To fix these we will pivot our data
##tidyr: pivot_longer()
##pivot_wider()
```

### 2.4 Lengthening datasets

pivot_longer() makes datasets “longer” by increasing the number of rows and decreasing the number of columns, solving those common problems of data values in the variable name.

```{r}
billboard

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )  

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
###There are now far fewer rows in total, indicating that a heap of NAs were dropped.

##Pivoting longer - example
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```

3 key arguments to the pivot_longer() function: - cols which specifies the columns you want to pivot (the ones that aren’t variables). Note: you could either use !c(artist, track, date.entered) OR starts_with('wk') because the cols argument uses the same syntax as select(). - names_to which names the variable stored in the column names. We chose to name that variable week. - values_to which names the variable stored in the cell values that we named rank

### 2.5 Widening datasets

```{r}
cms_patient_experience

cms_patient_experience |> 
  distinct(measure_cd, measure_title)

cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )

cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
####Pivoting wider - example
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

df |> 
  distinct(measurement) |> 
  pull()

df |> 
  select(-measurement, -value) |> 
  distinct()

df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
```

### 2.6 Pivoting Exercises using Palmer Penguins

```{r}
library(palmerpenguins)

# Create a long version of the penguins dataset

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)

penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )
#### Widening data with pivot_wider()
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

head(mass_matrix)
```

### 2.7 Separating and uniting data tables

```{r}
table3

table3 %>% 
  separate(rate, into = c("cases", "population"))

table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)

table5 %>% 
  unite(new, century, year, sep = "")
```

### 2.8 Wrangling strings and dates

```{r}
#### Standardising text with stringr
library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c("Nelly Bay", "nelly_bay", "NELLY BAY", "Geoffrey_Bay", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_"),
    # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean)
  )

print(clean_sites)


####Mastering time with lubridate
library(lubridate)
# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)
```

### 2.9 Relational data: Joining tables

```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "WP"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

#The workhorse: left_ and right_join()
# Joining metadata to our observations
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)

#The strict filter: inner_join()
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)

#The detective: anti_join()
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))
print(missing_context)

```

### 2.10 Handling missing values
```{r}
#Converting legacy errors with na_if()
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)

#Replacing missing values with coalesce()
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)

## Bonus tip: coalesce() is incredibly powerful. If you have two columns, like a primary_sensor and a backup_sensor, coalesce(primary_sensor, backup_sensor) will automatically use the primary reading, but seamlessly fill in the gaps with the backup reading if the primary had failed!

#The special case of NaN (Not a Number)
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc)

#Implicit missing values and the zero-data framework: complete()
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))
# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)

#The nuclear option: drop_na()
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity)
```

### 2.11 Practical Exercises: Penguins, Pivots, and Relational Data
```{r}
# Exercise 1: Cleaning the messy metadata

# Make sure packages are loaded
library(tidyverse)
library(palmerpenguins)

# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print("Original messy metadata:")
print(island_metadata)

# Create clean_metadata object
clean_metadata <- island_metadata %>%
  mutate(
    # Fix island_name: trim whitespace and convert to title case to match penguins dataset
    island_name = str_trim(island_name) %>% str_to_title(),
    # Convert station_install to proper Date object
    # Handle mixed date formats: some use "/" and some use "-"
    station_install = parse_date_time(station_install, orders = "dmy")
  )

print("\nCleaned metadata:")
print(clean_metadata)

# Verify it matches the penguins dataset islands
print("\nUnique islands in penguins dataset:")
print(unique(penguins$island))

print("\nUnique islands in clean_metadata:")
print(unique(clean_metadata$island_name))

```

```{r}
# Exercise 2: The Relational Join

# Make sure packages are loaded
library(tidyverse)
library(palmerpenguins)

# First, run Exercise 1 to create clean_metadata
# (Or copy the code from exercise_1_solution.R)

island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

clean_metadata <- island_metadata %>%
  mutate(
    island_name = str_trim(island_name) %>% str_to_title(),
    station_install = parse_date_time(station_install, orders = "dmy")
  )

# Create penguins_spatial by joining penguins with clean_metadata
penguins_spatial <- penguins %>%
  left_join(clean_metadata, by = join_by(island == island_name))

# Check that it worked - scroll to the right to see new columns
print("First few rows of penguins_spatial:")
head(penguins_spatial)

# You can also view the structure
print("\nStructure of penguins_spatial:")
glimpse(penguins_spatial)

# Verify the join worked by checking for NAs in the new columns
print("\nChecking new columns from metadata:")
print(penguins_spatial %>% 
  select(island, station_install, latitude) %>%
  head(10))

```

```{r}
# Exercise 3: The Wide Summary Matrix

# Make sure packages are loaded
library(tidyverse)
library(palmerpenguins)

# Assuming you have penguins_spatial from Exercise 2
# If not, run the previous exercises first

# Create the publication-ready summary table
body_mass_summary <- penguins_spatial %>%
  # Drop any missing values in the body_mass_g column
  drop_na(body_mass_g) %>%
  # Group by species and island
  group_by(species, island) %>%
  # Summarise to find the maximum body mass
  summarise(max_body_mass = max(body_mass_g), .groups = "drop") %>%
  # Pivot wider so species are rows and islands are columns
  pivot_wider(
    names_from = island,
    values_from = max_body_mass
  )

# Display the clean, publication-ready table
print("Maximum body mass (g) by species and island:")
print(body_mass_summary)

# Optional: Round to make it more publication-ready
body_mass_summary_rounded <- body_mass_summary %>%
  mutate(across(where(is.numeric), round, digits = 0))

print("\nRounded for publication:")
print(body_mass_summary_rounded)

```

###END
