---
title: "RforMarineScience_Week3_2.12"
format: html
editor: visual
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

#Introduction #Keystone 2.12 #House Keeping

```{r}
#Set options, housekeeping
knitr::opts_chunk$set(
  echo = FALSE, 
  include = TRUE, 
  message = FALSE, 
  warning = FALSE)
```

```{r housekeeping}
rm(list = ls())
```

#Packages

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readxl)
```

#Load files

```{r}
catch1 <- read_excel("C:/Users/andre/Techniques in Marine Science 1/Module_02/Module_02_RforMarineScience/data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("C:/Users/andre/Techniques in Marine Science 1/Module_02/Module_02_RforMarineScience/data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("C:/Users/andre/Techniques in Marine Science 1/Module_02/Module_02_RforMarineScience/data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("C:/Users/andre/Techniques in Marine Science 1/Module_02/Module_02_RforMarineScience/data/workshop2/estuary_catch_log.xlsx", sheet = 4)

catch_data <- bind_rows(catch1, catch2, catch3, catch4)
```

#loading all the data

```{r}
getwd()
metadata <- read_csv("C:/Users/andre/Techniques in Marine Science 1/Module_02/Module_02_RforMarineScience/data/workshop2/estuary_metadata.csv")
sensor_raw <- read_csv("C:/Users/andre/Techniques in Marine Science 1/Module_02/Module_02_RforMarineScience/data/workshop2/estuary_sonde_data.csv")
species_dict <- read_csv("C:/Users/andre/Techniques in Marine Science 1/Module_02/Module_02_RforMarineScience/data/workshop2/species_dictionary.csv")
```

#Bind row together

```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
rm(catch1, catch2, catch3, catch4) #clean up the environment by removing the individual data frames

```

#different way to look at the files

```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

#Can't tell from these outputs how many levels of site and species we have. Two options: let's check those explicitly by using distinct() or can convert to factors Let's look at the distinct values first, as we might need to clean before converting to factor vars.

```{r}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```

#Next step: #Fix site names and place in reasonable order. #Can decide this now or we can load the metadata and ensure consistency #Fix species names - to do this we'll need to make consistent and then use the taxonomic dictionary to translate to scientific names

#How to standardize the species names using the species dictionary

```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

#Recheck

```{r}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
glimpse(metadata)

```

# Check sensor data

```{r}
head(sensor_raw)
str(sensor_raw)
glimpse(sensor_raw)
summary(sensor_raw)
```

#### Notes:

- Environmental variables look like they are mostly in good shape (all numeric), however the turbidity data has a failed value (-999) so we'll need to deal with that
- Site looks ok, but again should check how many levels or distinct values we have:

###Clean up sensor data

```{r}
sensor_data <- sensor_raw |>
  mutate(
    #parse the messy character sting (day/month/year hour:minute)
    datetime = dmy_hm(timestamp),
    #Convert the -999.0 hardware error to a true NA
    clean_turbudity = na_if(turbidity, -999.0),
    #Convert negative salinity values to NA
    clean_salinity = if_else(salinity < 0, NA, salinity),
  )  
```

# Standardise variable in the catch log to match dictionaries

```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_"),
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

## Check common names in the catch log match the species dictionary

```{r}
catch_data |> 
  distinct(species)
species_dict |> 
  distinct(common_name)
```

# Taxonomic Translation (Common to Scientific)

```{r}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))
```

# Summarise 15-minute sonde data into daily means

```{r}
sensor_daily <- sensor_data |>
  mutate(
    date = floor_date(datetime, "day"),
    date = as_date(date) # Ensure it matches the Date format of the catch log
  ) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(clean_turbudity, na.rm = TRUE),
    .groups = "drop"
  )
```

# Implicit zero observations

4 x ssp 4 x sites 30 surveys 4*4*30 = 480 possible observations

# Now we need to add in the implicit zero observations for species that were not caught in a given survey. To do this, we can create a complete grid of all possible combinations of site, date, and species, and then join this with our existing catch data.

```{r}
# 1. Join metadata and sensor data
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))

# 2. Implicit zero observations
catch_data_final<-
catch_data_final|>
  complete(
    nesting(site, date,lat, lon, zone, mean_temp, mean_salinity, mean_turbidity),
  scientific_name
  ) |>
  mutate(
  #Cinvert the resulting NA counts to 0
  count = coalesce(count, 0) 
  )

```

# Check that there is an actual difference in salinity between zones: 
```{r}
sal_sum <- 
 catch_data_final |>
  group_by(zone) |>
  summarise(
    sem_salinity = sd(mean_salinity) / sqrt(n()),
    #sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )
```

# Make point and error plot
```{r}
sal_sum|>
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sem_salinity, ymax = median_salinity + sem_salinity), width = 0.2) +
  labs(title = "Median Salinity by Zone with SEM Error Bars", x = "Zone", y = "Median Salinity") +
  theme_minimal()
```
#Now calculate 
```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count, na.rm = TRUE),
    sem_catch = sd(count, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )
```

#Graph it

```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_point(position = "dodge", size = 2, pch = 23) +
  #geom_line(group = 1) +
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) +
  # add in points from master data
  # geom_point(data = catch_data_final, aes(x = zone, y = count), 
  #           position = position_jitter(width = 0.2), alpha = 0.5, size = 1, inherit.aes = FALSE) +
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"),
   # legend.position = "bottom"
    legend.position = "none"
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients",
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```


